---
title: "Australia's exports of endangered species"
author: "Rosa Mar Dominguez Martinez"
date: "2024-01-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries

```{r libraries}
library(tidyverse)
library(here) 
library(janitor)
library(readxl)
library(ggbreak)
library(RColorBrewer)
library(ggsankey)
library(scales) ## to show palettes, including hex values in the colors
library(ggalluvial)
library(ggpubr)
library(ggthemes)
here()
```

## Import data


```{r data, echo=FALSE}
trade <- read_csv(here("_input_data/20231113_australia_ARTIS_australia_cites_iucn.csv")) # not publicly available

# production <- read_csv(here("_input_data", "FAO_fishstat/production_isscaap_species.csv")) |> 
#   clean_names() |> 
#   select(-unit) |>
#   rename(country = country_name,
#          sp_name = asfis_species_name_2, 
#          isscaap_group = asfis_species_name_3,
#          #isscaap_group = asfis_species_name_1,
#          fishing_area = fao_major_fishing_area_name,
#          # fishing_area_code = fao_major_fishing_area_code, 
#          unit = unit_name ) |> 
#   filter(country == "Australia") |> 
#   pivot_longer(cols = -c(1:7), names_to = "year", values_to = "value") |> 
#   mutate(year = gsub(pattern = "x", replacement = "", year) |> as.numeric()) 


epbc <-  read_excel(here("_input_data/EPBC/EPBC_fauna_2023.xls")) |> clean_names() |> 
  mutate(species = tolower(species)) 
```



```{r, join}

#join trade with EPBC data

trade |> filter(source_country_iso3c == "AUS") 
  
(trade_epbc <- trade |> 
  filter(source_country_iso3c == "AUS") |> 
  left_join(epbc, by = c("sciname" = "species")) |> 
  mutate(listed = case_when(!is.na(common_name) ~ "listed", 
                            T ~ "not_listed")))
#species listed and exported 
# sciname                
#   <chr>                  
# 1 bidyanus bidyanus      
# 2 hoplostethus atlanticus
# 3 maccullochella peelii  
# 4 thunnus maccoyii       
# 5 galeorhinus galeus     
# 6 seriolella brama    

#because bidyanus bidyanus (silver perch) and maccullochella peelii (Murray cod) are farmed, exclude them

(trade_epbc <- trade |> 
  filter(source_country_iso3c == "AUS") |> 
  left_join(epbc, by = c("sciname" = "species")) |>
  mutate(listed = case_when(sciname %in% c("galeorhinus galeus", "hoplostethus atlanticus", 
                                           "thunnus maccoyii", "seriolella brama") ~ "listed", 
                            T ~ "not_listed")) )
```
How much exports of EPBC - lisdted species since they were included in the Act
```{r stats}

#trade after each species was listed 
### total volumes since they were listed 
(listed_total <- trade_epbc |> filter(listed == "listed") |>
  mutate(bef_af = case_when(sciname == "hoplostethus atlanticus" & year >2006 ~ "after",
                            sciname == "galeorhinus galeus" & year > 2009 ~ "after", 
                            sciname == "seriolella brama" & year >2015 ~ "after", 
                            sciname == "thunnus maccoyii" & year >2010 ~ "after", 
                            T ~ "not_listed")))

#total volume of each species traded before and after they were listed
listed_total |> 
    group_by(sciname, bef_af) |> 
    reframe(total = sum(live_weight_t))

listed_total |> group_by(sciname, bef_af) |> 
    reframe(total = sum(live_weight_t)) |> filter(bef_af == "after") |> reframe(tot = sum(total)) 
# 106615.7 tonnes of all listed species traded since they were listed 

listed_total |> filter(bef_af == "after", exporter_iso3c != "AUS") |>  select(exporter_iso3c) |> distinct()
#39 countries
listed_total |> filter(bef_af == "after", importer_iso3c != "AUS") |> select(importer_iso3c) |> distinct()
#97 countries

#Australia exports EPBC-listed species to 136 countries

# % of exports of tuna 
listed_total |> mutate(species = case_when(sciname == "thunnus maccoyii" ~ "tuna", 
                                           T ~ "Other")) |> 
  group_by(species) |> 
  reframe(total = sum(live_weight_t), live_weight_t) |> 
  select(-live_weight_t) |> distinct() 

224449.21*100/(21667.55 + 224449.21)


```

transform data so there is no intermediary countries 
```{r exporter is importer}
#the ones imported from intermediaries 
(importer_1<- listed_total |> 
  filter(source_country_iso3c != exporter_iso3c) |> 
  select(-importer_iso3c) |> 
  rename(importer_iso3c = exporter_iso3c))

#processed by australia
(importer_2<- listed_total |> 
  filter(source_country_iso3c == exporter_iso3c) |> 
  select(- source_country_iso3c) |> 
  rename(source_country_iso3c = exporter_iso3c))

(no_inter <- rbind(importer_1, importer_2))
```

line fig with exports of each listed species through time
```{r fig 1}
# 
# ggplot(no_inter |> filter(listed == "listed") |> mutate(common_name = case_when(sciname =="hoplostethus atlanticus" ~ "Orange Roughy", 
#                                                                                   sciname == "galeorhinus galeus" ~ "School Shark", 
#                                                                                   T ~ common_name)) |> 
#          group_by(common_name, year) |> reframe(total = sum(live_weight_t))) + 
#   geom_line(aes(x = year, y = total, color = common_name), linewidth = 1) + 
#    scale_x_discrete(limits = c(seq(1996,2020, 5), 2020)) +
#   scale_y_break(c(4000,6000)) +
#    #xlim(1996, 2020) +
#   #scale_y_break(c(800, 1000)) +
#   scale_color_brewer(palette = "Spectral") +
#   geom_segment(aes(x = 2000, y = -0.1, xend = 2000, yend = 12000), linewidth = 1) + #,arrow = arrow(length = unit(2, "mm"))) +  
#   annotate("text", x = 2000, y = 12500, label = "implementation\nof EPBC Act") + 
#   geom_segment(aes(x = 2006, y = -0.1, xend = 2006, yend = 3200), color = "grey", alpha =0.4, linetype = "dashed", linewidth = 1) +  
#   #annotate(geom = "point", x = 2006, y = 4500) + 
#   annotate("text", x = 2006, y = 3500, label = "Orange\nRoughy") + 
#   geom_segment(aes(x = 2009, y = -0.1, xend = 2009, yend = 3000), color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 0.9) +  
#   #annotate(geom = "point", x = 2009, y = 5500) + 
#   annotate("text", x = 2009, y = 3300, label = "School\nShark") + 
#   geom_segment(aes(x = 2015, y = -0.1, xend = 2015, yend = 3200), color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 1) +  
#   #annotate(geom = "point", x = 2015, y = 4500) + 
#   annotate("text", x = 2015, y = 3500, label = "Blue\nWarehou") +
#   geom_segment(aes(x = 2010, y = -0.1, xend = 2010, yend = 11000),color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 1) +  
#   #annotate(geom = "point", x = 2010, y = 11500) + 
#   annotate("text", x = 2010, y = 11300, label = "Southern\nBluefin Tuna") + 
#   theme_pubr( base_size = 14,legend = "right") + 
#   # theme(axis.text.x = element_text(angle = 60, hjust=1), 
#   #       text = element_text(size=16),
#   #       axis.title = element_text(size = 14), 
#   #       axis.text = element_text(size = 14),
#   #       legend.text = element_text(size=14),
#   #       legend.key.size = unit(0.6, "cm"), 
#   #       plot.margin = margin(1,1,1,1, "cm")) + 
#   labs(y = "Exports live weight equivalent (T)", color = "")

#ggsave("fig1_export_time_v2.pdf", device = "pdf", path = here("AusTrade_ThrSpp/figures/exports"), width = 38, height = 20, units = c("cm"), dpi = 600 )

```

line fig with exports of each listed species through time - square root
```{r fig 1}

ggplot(no_inter |> filter(listed == "listed") |> mutate(common_name = case_when(sciname =="hoplostethus atlanticus" ~ "Orange roughy", 
                                                                                  sciname == "galeorhinus galeus" ~ "School shark", sciname == "thunnus maccoyii" ~ "Southern bluefin tuna", 
                                                                                 sciname == "seriolella brama" ~ "Blue warehou", 
                                                                                
                                                                                  T ~ common_name)) |> 
         group_by(common_name, year) |> reframe(total = sum(live_weight_t))) + 
  geom_line(aes(x = year, y = sqrt(total), color = common_name), linewidth = 1) + 
    scale_x_discrete(limits = c(seq(1996,2020, 5), 2020)) +
   scale_y_discrete(limits = c(seq(0,110, 10))) +
  # scale_y_break(c(4000,6000)) +
  #  #xlim(1996, 2020) +
  #scale_y_break(c(800, 1000)) +
  scale_color_brewer(palette = "Spectral") +
  geom_segment(aes(x = 2000, y = -0.1, xend = 2000, yend = 99), linewidth = 1) + #,arrow = arrow(length = unit(2, "mm"))) +  
  annotate("text", x = 2000, y = 102, label = "implementation\nof EPBC Act") + 
  geom_segment(aes(x = 2006, y = -0.1, xend = 2006, yend =30), color = "grey", alpha =0.4, linetype = "dashed", linewidth = 1) +  
  #annotate(geom = "point", x = 2006, y = 4500) + 
  annotate("text", x = 2006, y = 34, label = "Orange\nroughy") + 
  geom_segment(aes(x = 2009, y = -0.1, xend = 2009, yend = 28), color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 0.9) +  
  #annotate(geom = "point", x = 2009, y = 5500) + 
  annotate("text", x = 2009, y = 30, label = "School\nshark") + 
  geom_segment(aes(x = 2010, y = -0.1, xend = 2010, yend = 102),color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 1) +  
  #annotate(geom = "point", x = 2010, y = 11500) + 
  annotate("text", x = 2010, y = 105, label = "Southern\nbluefin tuna") +
  geom_segment(aes(x = 2015, y = -0.1, xend = 2015, yend = 30), color = "grey", alpha = 0.4, linetype = "dashed", linewidth = 1) +  
  #annotate(geom = "point", x = 2015, y = 4500) + 
  annotate("text", x = 2015, y = 33, label = "Blue\nwarehou") +
  theme_pubr(base_size = 14,legend = "right") + 
  # theme(axis.text.x = element_text(angle = 60, hjust=1), 
  #       text = element_text(size=16),
  #       axis.title = element_text(size = 14), 
  #       axis.text = element_text(size = 14),
  #       legend.text = element_text(size=14),
  #       legend.key.size = unit(0.6, "cm"), 
  #       plot.margin = margin(1,1,1,1, "cm")) + 
  labs(y = "Exports live weight equivalent (sqrt(T))", color = "")

#ggsave("fig1_sqrt_export_time.pdf", device = "pdf", path = here("AusTrade_ThrSpp/figures/exports"), width = 38, height = 20, units = c("cm"), dpi = 600 )

```


trade flow data 
change data format
```{r flow plot}
(epbc_trade_flow <- trade_epbc |> filter(sciname %in% c("hoplostethus atlanticus", "seriolella brama", "galeorhinus galeus")) |> 
  mutate(bef_af = case_when(sciname == "hoplostethus atlanticus" & year >2006 ~ "after",
                            sciname == "galeorhinus galeus" & year > 2009 ~ "after", 
                            sciname == "seriolella brama" & year >2015 ~ "after", 
                            sciname == "thunnus maccoyii" & year >2010 ~ "after", 
                            T ~ "not_listed")))
 
flow_df_orange<- epbc_trade_flow |> 
  filter(sciname == "hoplostethus atlanticus", bef_af == "after") |> filter_artis(prod_method = "capture") |> 
  mutate(sciname = "hoplostethus atlanticus")

flow_df_blue<- epbc_trade_flow |> 
  filter(sciname == "seriolella brama", bef_af == "after") |> filter_artis(prod_method = "capture") |> 
  mutate(sciname = "seriolella brama")

flow_df_school<- epbc_trade_flow |> 
  filter(sciname == "galeorhinus galeus", bef_af == "after") |> filter_artis(prod_method = "capture") |> 
  mutate(sciname = "galeorhinus galeus")

(flow_df <- rbind(flow_df_orange, flow_df_blue, flow_df_school))
 
```

fig 2, trade flow alluvial plot version
```{r}
flow_df

## try alluvial plot
(flow_aus<- flow_df |> filter(producers == exporters & producers == "AUS") |> 
  select(exporters, importers, sciname, total_q) |> 
  group_by(sciname,exporters, importers) |> 
  reframe(total = sum(total_q)) |> 
  rename(producers = exporters))

(flow_else <- flow_df |> filter(producers != exporters) |> 
  select(producers,exporters, sciname, total_q) |> 
  group_by(sciname,exporters, producers) |> 
  reframe(total = sum(total_q)) |> 
  rename(importers = exporters))

all_flow<- rbind(flow_aus, flow_else)

ggplot(all_flow, aes(axis1 = sciname, axis2 = importers, y = total)) + 
  geom_flow(aes(fill = sciname, color = sciname), width = 1.95/16, knot.pos = 0.5, alpha = 0.6) +
        geom_stratum(width = 1.95/16, , alpha = 0.4,  color = "black") +
  #geom_alluvium(aes(fill = sciname), curve_type = "arctangent") + 
  #geom_stratum() + 
  geom_text(stat = "stratum", aes(label =after_stat(stratum))) + 
  #scale_fill_hue(labels = c("Galeorhinus galeus", "Hoplostethus atlanticus", "Seriolella brama")) + 
  scale_fill_viridis_d() + 
  scale_color_viridis_d() +
  scale_x_discrete(limits = c("exported taxa", "importers"),
                   expand = c(0.18, 0.08)) +
  scale_y_continuous(breaks = NULL)+
  theme_hc() + 
  guides(fill = guide_legend(title = "Exported taxa")) + 
  theme(axis.text.x = element_text(vjust = 7, face="bold", size=12, color="black"), 
        #axis.title.x = element_text(face="bold", size=16, color="black"),
        axis.text.y = element_text(face="bold", size=12, color="black"),
        axis.title.y = element_blank(),
        strip.text = element_text(size=10, face="bold"),
        plot.title = element_text(size=20, face="bold"),
        legend.position = "none")

#ggsave("alluvial_flow.png", device = "png", path = "AusTrade_ThrSpp/figures/exports/", width = 50, height = 30, units = c("cm"), dpi = 400 )
```

Species IUCN and EPBC
```{r, table}

# table with sciname, common name, EPBC listing, IUCN listing, year of listing, volume

table_dat_epbc <- trade_epbc |> filter(listed == "listed") |> group_by(sciname, year) |> 
  mutate(bef_af = case_when(sciname == "hoplostethus atlanticus" & year >2006 ~ "after",
                            sciname == "galeorhinus galeus" & year > 2009 ~ "after", 
                            sciname == "seriolella brama" & year >2015 ~ "after", 
                            sciname == "thunnus maccoyii" & year >2010 ~ "after", 
                            T ~ "before"), 
         iucn_epbc = "EPBC", 
         listing_date = case_when(sciname == "hoplostethus atlanticus" ~ 2006,
                            sciname == "galeorhinus galeus"~ 2009, 
                            sciname == "seriolella brama" ~2015, 
                            sciname == "thunnus maccoyii" ~2010, 
                            T ~ NA)) |> 
  group_by(sciname, bef_af) |> 
  reframe(total_exported = sum(live_weight_t), common_name, listing_date, iucn_epbc, iucn_redlist_category) |> unique()


table_dat_iucn <- trade_epbc |> 
  filter(method == "capture", 
         iucn_redlist_category %in% c("Critically Endangered", "Endangered", "Near Threatened", "Vulnerable"), 
         sciname != c("hoplostethus atlanticus", "galeorhinus galeus", "seriolella brama","thunnus maccoyii")) |> 
  mutate(listing_date = case_when(sciname == "galeorhinus galeus" ~ 2000, 
                                  sciname == "thunnus maccoyii" ~ 1996, 
                                  sciname == "anguilla australis" ~ 2019, 
                                  sciname == "atractoscion aequidens" ~ 2020, 
                                  sciname == "bathyraja irrasa" ~ 2009, 
                                  sciname == "bidyanus bidyanus" ~ 1996, 
                                  sciname == "carcharhinus amboinensis" ~ 2000,    ## DD before 2021
                                  sciname == "carcharhinus sorrah" ~ 2009,
                                  sciname == "cherax destructor" ~ 1996,  #not assessed since
                                  sciname == "eusphyra blochii" ~ 2003, 
                                  sciname == "haliotis rubra" ~ 2021, 
                                  sciname == "istiophorus platypterus" ~ 2022, # 2011 least concern 
                                  sciname == "lamna nasus" ~ 1996, 
                                  sciname == "makaira nigricans" ~ 2011, 
                                  sciname == "pomatomus saltatrix" ~ 2015, 
                                  sciname == "rhizoprionodon acutus" ~ 2020, #2003 LC, 
                                  sciname == "scomberomorus commerson" ~ 2011,
                                  sciname == "somniosus pacificus" ~ 2009, #DD before 2021
                                  sciname == "thunnus obesus" ~ 1996, 
                                  sciname == "xiphias gladius" ~ 1996, 
                                  T ~ NA),                           
         common_name = case_when(sciname == "anguilla australis" ~ "shortfin eel", 
                                 sciname == "atractoscion aequidens" ~ "african weakfish", 
                                 sciname == "bathyraja irrasa" ~ "kerguelen sandpaper skate", 
                                 sciname == "bidyanus bidyanus" ~ "silver perch", 
                                 sciname == "carcharhinus amboinensis" ~ "pigeye shark",
                                 sciname == "carcharhinus sorrah" ~ "spottal shark",
                                 sciname == "eusphyra blochii" ~ "winghead shark",
                                 sciname == "haliotis rubra" ~ "blacklip abalone", 
                                 sciname == "istiophorus platypterus" ~ "sailfish", 
                                 sciname == "lamna nasus" ~ "porbeagle", 
                                 sciname == "makaira nigricans" ~ "blue marlin", 
                                 sciname == "pomatomus saltatrix" ~ "blue fish", 
                                 sciname == "rhizoprionodon acutus" ~ "milk shark", 
                                 sciname == "scomberomorus commerson" ~ "narrow-barred spanish mackerel", 
                                 sciname == "somniosus pacificus" ~ "pacific sleeper shark", 
                                 sciname == "thunnus obesus" ~ "bigeye tuna",
                                 sciname == "thunnus maccoyii" ~ "southern bluefin tuna",
                                 sciname == "xiphias gladius" ~ "swordfish", 
                                 sciname == "galeorhinus galeus" ~ "school shark",
                                 T ~ sciname), 
         iucn_epbc = "IUCN",
         bef_af = case_when(year > listing_date ~ "after", 
                            T ~ "before")) |> 
  group_by(sciname, bef_af) |>
  reframe(total_exported = sum(live_weight_t), common_name, listing_date, iucn_epbc, iucn_redlist_category) |> unique()

table_listed <-table_dat_epbc |> 
  rbind(table_dat_iucn) |> distinct()


#write.csv(table_listed, here("tables/exported_vols_iucn-epbc.csv"), row.names = FALSE)

## porbeagle 

porbeagle<- trade_epbc |> filter(grepl("lamna", sciname)) |> mutate(bef_af = case_when(year > 2009 ~ "after", 
                                                                                       T~ "before")) |> 
  group_by(year, bef_af) |> 
  reframe(total = sum(live_weight_t))



```

Other species nominated for EPBC 
```{r, nominated sp}

#### species proposed####

not_assessed <-read_excel(here("_input_data/EPBC/EPBC_nominated_sp_11Dec2023.xlsx"), sheet = 1) |> 
  clean_names() |> mutate(sciname = tolower(species_name)) |> 
  select(-species_name)

not_assessed_trade <- not_assessed |> left_join(trade_epbc, by = "sciname", relationship = "many-to-many") |> 
  filter(!is.na(source_country_iso3c))

assessed <- read_excel(here("_input_data/EPBC/EPBC_nominated_sp_11Dec2023.xlsx"), sheet = 2) |> 
  clean_names() |> mutate(sciname = tolower(sciname))

assessed_trade <- assessed |> left_join(trade_epbc, by = "sciname", relationship = "many-to-many") |> 
  filter(!is.na(source_country_iso3c))

```

```{r, CITES}

# check with CITES 

CITES <- read_csv(here("_input_data/Index_of_CITES_Species_2024-01-03 05 25.csv")) |> 
  filter(Kingdom == "Animalia") |> 
  mutate(sciname = tolower(FullName)) |> 
    select(sciname, CurrentListing, CitesAccepted)

CITES_epbc <- trade_epbc |> 
  left_join(CITES, by = "sciname") 

CITES_epbc |> filter(!is.na(CurrentListing)) |> count(sciname, CurrentListing, year)


#none of these species were traded before they were listed 
#https://checklist.cites.org/#/en/search/output_layout=alphabetical&level_of_listing=0&show_synonyms=1&show_author=1&show_english=1&show_spanish=1&show_french=1&scientific_name=&page=1&per_page=20
```